---
title: "Statistics and Data Analysis Using R Programming"
author: "Isaac Ajao"
date: "2025-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### **1. Introduction to R for Statistics and Data Analysis**  
- Overview of R and its importance in statistics & data analysis  
- Installing R and RStudio  
- Basic R syntax (variables, data types, functions)  
- Loading essential libraries (`tidyverse`, `ggplot2`, `dplyr`)  

### **2. Core Statistical Techniques Using R**  
- **Descriptive Statistics**: Mean, median, mode, variance, standard deviation (`summary()`, `sd()`, `IQR()`)  
- **Data Visualization**: Histogram, boxplot, scatterplot (`ggplot2`)  
- **Hypothesis Testing**: t-tests, ANOVA (`t.test()`, `aov()`)  
- **Regression Analysis**: Simple & multiple linear regression (`lm()`, `summary()`)  

### **3. Data Analysis Workflow**  
- **Importing Data**: CSV, Excel, database connections (`read.csv()`, `readxl`)  
- **Data Cleaning**: Handling missing values (`na.omit()`, `tidyverse` functions)  
- **Data Transformation**: Filtering, selecting, mutating columns (`dplyr::filter()`, `mutate()`)  
- **Exploratory Data Analysis (EDA)**: Summarizing and visualizing key patterns  

### **4. Real-World Use Case: Data-Driven Decision Making**  
- Present a case study (e.g., analyzing customer satisfaction, sales trends, environmental data)  
- Walk through the full process: data import → cleaning → analysis → visualization → interpretation  
- Discussion on insights and how to communicate findings effectively  

---

### **1. Introduction to R for Statistics and Data Analysis**  

#### **R Script**: Basic operations  
```{r}
# Install and load essential libraries
# install.packages("tidyverse")
library(tidyverse)

# Basic operations
x <- c(10, 20, 30, 40, 50)
mean(x)  # Compute mean
sd(x)    # Compute standard deviation
summary(x)  # Get a statistical summary
```

#### **Exercise**:  
1. Install R and RStudio.  
2. Load the `tidyverse` package.  
3. Create a numeric vector of any five values and compute the mean and standard deviation.  

### **2. Core Statistical Techniques Using R**  

#### **Students' Scores Dataset**  
```{r}
# Create Students' Scores Dataset
students_scores <- data.frame(
  Name = c("Ade", "Chuks", "Bayo", "Mary", "Afusat"),
  Score = c(85, 78, 90, 70, 88),
  Study_Hours = c(10, 8, 12, 6, 11)
)

print(students_scores)
```

#### **Exercise**:  
1. Check for missing values
1. Calculate mean, median, and standard deviation of scores.  
2. Perform a t-test to check if the mean score is different from 75.  
3. Plot a histogram of student scores.  
4. Run a linear regression model to predict scores based on study hours. 


#### **R Script**:  
```{r}
# Load the dataset
students <- students_scores

# Descriptive statistics
summary(students)

# Histogram of scores
library(ggplot2)
ggplot(students, aes(x = Score)) + geom_histogram(binwidth = 5, fill = "blue", color = "black")

# Hypothesis test (t-test)
t.test(students$Score, mu = 75)  # Checking if the average score is significantly different from 75

# Linear regression: Study Hours vs Score
model <- lm(Score ~ Study_Hours, data = students)
summary(model)
```


### **Using large data from Csv/Excel file**

#### **R Script**:  

#### **Check for missing values and distribution**
```{r}
# Import the dataset
students <- read.csv("students_scores_nigeria_1000.csv")

# Check for missing values
sum(is.na(students))

# Check distribution of Scores using boxplot
ggplot(students, aes(y = Score)) + geom_boxplot(fill="skyblue", color="black")

# Pairwise correlation analysis
cor(students[, c("Score", "Study_Hours")])
```

#### **Creating new variables**
```{r}
# Categorize students based on performance
students <- students %>%
  mutate(Performance_Category = case_when(
    Score >= 85 ~ "Excellent",
    Score >= 70 & Score < 85 ~ "Good",
    TRUE ~ "Needs Improvement"
  ))

# View the first few rows
head(students)

```


```{r}
# Descriptive statistics
summary(students)

# Histogram of scores
library(ggplot2)
ggplot(students, aes(x = Score)) + geom_histogram(binwidth = 5, fill = "blue", color = "black")

# Hypothesis test (t-test)
t.test(students$Score, mu = 75)  # Checking if the average score is significantly different from 75

# Linear regression: Study Hours vs Score
model <- lm(Score ~ Study_Hours, data = students)
summary(model)
```

### **3. Data Analysis Workflow**  

#### **Sales Data Dataset**  
```{r}
# Create Sales Data Dataset
sales_data <- data.frame(
  Date = as.character(seq(as.Date("2024-01-01"), by = "day", length.out = 5)),
  Sales = c(500, 700, 800, 400, 650),
  Category = c("Electronics", "Clothing", "Electronics", "Clothing", "Electronics")
)

print(sales_data)
```  

#### **Exercise**:  
1. Import `sales_data.csv` into R.  
2. Convert the Date column to the correct format.  
3. Filter only Electronics sales and compute the total revenue.  
4. Create a time series line plot showing sales trends by category.

#### **R Script**: 
#### **ANOVA test to compare sales across multiple categories**
```{r}
# ANOVA to compare sales among different product categories
anova_model <- aov(Sales ~ Category, data = sales_data)
summary(anova_model)

```


```{r}
# Load dataset
sales <- read.csv("sales_data.csv")

# Convert Date column to Date format
sales$Date <- as.Date(sales$Date, format="%Y-%m-%d")

summary(sales) # Summary statistics

# Filter sales for Electronics only
library(dplyr)
electronics_sales <- filter(sales, Category == "Electronics")

# Aggregate total sales by category
total_sales <- sales %>% group_by(Category) %>% 
summarise(Total = sum(Sales))
print(total_sales)

# Plot time series sales trend
ggplot(sales, aes(x = Date, y = Sales, color = Category)) + geom_line() + geom_point()
```

### **Using large data from Csv/Excel file**

#### **R Script**: 

```{r}
# Import dataset
sales <- read.csv("sales_data_1000.csv")

# Convert Date column to Date format
sales$Date <- as.Date(sales$Date, format="%Y-%m-%d")

summary(sales) # Summary statistics

# Filter sales for Electronics only
library(dplyr)
electronics_sales <- filter(sales, Category == "Electronics")

# Aggregate total sales by category
total_sales <- sales %>% group_by(Category) %>% 
summarise(Total = sum(Sales))
print(total_sales)

# Plot time series sales trend
ggplot(sales, aes(x = Date, y = Sales, color = Category)) + geom_line() + geom_point()
```



### **4. Real-World Use Case: Data-Driven Decision Making**  

#### **Employee Performance Dataset**  
```{r}
# Create Employee Performance Dataset
employee_performance <- data.frame(
  Employee = c("A", "B", "C", "D", "E"),
  Work_Hours = c(35, 40, 45, 50, 38),
  Performance_Score = c(78, 85, 90, 92, 80)
)

print(employee_performance)
```

#### **Exercise**:  
1. Load the `employee_performance.csv` dataset.  
2. Create a scatter plot with a trendline showing the relationship between work hours and performance.  
3. Run a linear regression to analyze how work hours impact performance.  
4. Predict performance for an employee working 42 hours.  

### **Using large data from Csv/Excel file**

#### **R Script**:  
```{r}
# Load dataset
performance <- read.csv("employee_performance.csv")

# Scatter plot
ggplot(performance, aes(x = Work_Hours, y = Performance_Score)) + geom_point() + geom_smooth(method="lm")

# Linear regression model
perf_model <- lm(Performance_Score ~ Work_Hours, data = performance)
summary(perf_model)

# Predict performance for a new employee working 42 hours
new_data <- data.frame(Work_Hours = 42)
predict(perf_model, new_data)
```


#### **R Script**:  
```{r}
# Load dataset
performance <- read.csv("employee_performance_1000.csv")

# Scatter plot
ggplot(performance, aes(x = Work_Hours, y = Performance_Score)) + geom_point() + geom_smooth(method="lm")

# Linear regression model
perf_model <- lm(Performance_Score ~ Work_Hours, data = performance)
summary(perf_model)

# Predict performance for a new employee working 42 hours
new_data <- data.frame(Work_Hours = 42)
predict(perf_model, new_data)
```
```

